The role of cross-linguistic lexical similarity on bilingual word acquisition

### 2021-01-15 Rmd/trajectories_manuscript.tex
Separate vocab type

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The role of cross-linguistic lexical similarity on bilingual word acquisition},
  pdfauthor={Gonzalo García-Castro1, Daniela Avila-Varela1, \& Núria Sebastian-Galles1},
  pdflang={en-EN},
  pdfkeywords={lexical acquisition, vocabulary, bilingualism},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\title{The role of cross-linguistic lexical similarity on bilingual word acquisition}
\author{Gonzalo García-Castro\textsuperscript{1}, Daniela Avila-Varela\textsuperscript{1}, \& Núria Sebastian-Galles\textsuperscript{1}}
\date{}

@@

Correspondence concerning this article should be addressed to Gonzalo García-Castro, Ramon Trias Fargas, 25-27, 08005 Barcelona, Spain. E-mail: \href{mailto:gonzalo.garciadecastro@upf.edu}{\nolinkurl{gonzalo.garciadecastro@upf.edu}}

@@

The structure of the lexicon also impacts the order of acquisition of new words. Words that share a high degree of phonological and/or semantic overlap with words already acquired are more likely to be acquired next ({\textbf{???}}). ({\textbf{???}}) and ({\textbf{???}}) showed that the connectivity of a given semantic category (i.e., animals) in a child's lexicon predicts a better performance in a disambiguation task, where participants are presented with a novel label in the presence of a familiar and a novel object: infants show stronger looking preference for the novel object if it belonged to a category for which many words had already been acquired. This points to the structure of the lexicon facilitating the strategies children engage during word learning.

@@

When compared to their monolingual peers, they know fewer words only when just on language is considered (e.g., English monolinguals know more words in English than English-Spanish bilinguals). When both languages are taken into account, bilingual children seem to know, at least, as many words as monolinguals do (Ben-Zeev, 1977; Bialystok, Luk, Peets, \& Yang, 2010; Blom et al., 2019; Core, Hoff, Rumiche, \& Señor, 2013; Doyle \& others, 1977; Fernandez, Pearson, Umbel, Oiler, \& Molinet-Molina, 1992; Hoff et al., 2012; Hoff, Rumiche, Burridge, Ribot, \& Welsh, 2014; Pearson \& Fernández, 1994; Rosenblum \& Pinker, 1983; but see Pearson, Fernández, \& Oller, 1993; Houwer, Bornstein, \& Putnick, 2014).

Floccia et al. (2018), gathered parental vocabulary estimates ({\textbf{???}}) from a large sample of 24-month-old toddlers learning English and an additional language (from a pool of 13 linguistically diverse languages). They computed the average phonological similarity between translation equivalents (TEs from now on; e.g., \emph{table} in English and \emph{tafel} in Dutch) for each pair of languages, and found a positive association between this measure of phonological similarity and participants' productive vocabulary sizes in their additional language. Using a similar measure of lexical similarity, Blom et al. (2019) extended these results to bilingual children aged three to 10, and reported a positive association between the lexical distance and comprehensive vocabulary size: bilinguals learning lexically close languages showed similar vocabulary sizes to those of monolinguals, but those learning the more distant languages showed lower vocabulary sizes.

These results suggest that linguistic distance between languages plays an important role during bilingual lexical acquisition, and that this effect operates within TEs. What mechanisms underlie this effect, and why they seem to operate differently across comprehension and production, or across the dominant and the non-dominant language, remain open issues. Floccia et al. (2018) pointed to parallel activation as a candidate mechanism behind this facilitation effect. The parallel activation principle suggests that bilinguals activate lexical representations in a language non-selective way: during the comprehension or production of a word in one language (e.g., \emph{cat}, in English) its corresponding translation in the other language is activated too {[}e.g., \emph{gato}, in Spanish. There is a vast body of evidence supporting language non-selective lexical access in both adults (e.g., Dell \& O'Seaghdha, 1991; {\textbf{???}}; Bobb, Von Holzen, Mayor, Mani, \& Carreiras, 2020; Dijkstra, 2005; Hoshino \& Kroll, 2008; Kroll, Gullifer, \& Rossi, 2013; Singh, 2014; Spivey \& Marian, 1999; Thierry \& Wu, 2007; Von Holzen, Fennell, \& Mani, 2019; Yudes, Macizo, \& Bajo, 2010; see Costa, Santesteban, \& Caño, 2005 for review), and children (e.g.~Von Holzen \& Mani, 2012; Bosma \& Nota, 2020; Jardak \& Byers-Heinlein, 2019; Poulin-Dubois, Bialystok, Blaye, Polonia, \& Yott, 2013). A critical implication of parallel activation is that the word form (i.e., phonology, orthography, etc.) of both translation equivalents are available to each other, impacting the comprehension or production dynamic of any of them.

@@

Under this account, cognates (i.e., phonologically similar TEs) would be acquired earlier than non-cognates, but this effect would only play a role once one of the members of the TE has been acquired. The evidence supporting an earlier age of acquisition of cognates relative to non-cognates if sparse. Schelletter (2002)`s reported a longitudinal case-study of a German/English child who produced cognate TEs earlier than non-cognate TE. Bosch and Ramon-Casas (2014), showed converging results from a sample of 48 monolingual and bilingual infants aged 24 months. The low number of words included in the analyses of both studies, and the lack of adjustment for lexical frequency, limits severely the strength of any conclusions concerning the difference in age of acquisition of cognates and non-cognates. On the other hand Floccia et al. (2018) worked with aggregated estimates of participants' vocabulary size, losing item-level information that could provide information about change in age of acquisition of TEs associated with their degree of phonological similarity. Therefore, their results can only be considered as compatible with cognateness playing a role on age of acquisition. In this study, we aimed at overcoming these pitfalls by using unaggregated data from individual responses, testing the role of cognateness explicitly and adjusting for lexical frequency.

@@

We implemented an on-line questionnaire using \texttt{formr} (Arslan, Walther, \& Tata, 2020), divided in three forms: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory (Hamilton, Plunkett, \& Schafer, 2000) and consisted in two lists of words, one in Catalan and one in Spanish). Items in one language were translation equivalents of the items in the other (e.g., whenever \emph{gos} {[}dog{]} was included in the Catalan inventory, the word \emph{perro} was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both (e.g., Catalan \emph{acabar} {[}\emph{to finish}{]} and Spanish \emph{acabar} and \emph{terminar}), or merged them into the same one (e.g., Spanish \emph{mono} {[}\emph{monkey}{]} and Catalan \emph{mono/mico}). We included items from a diverse sample of 26 semantic/functional categories (see Appendix 1). We discarded the following categories: adverbs, auxiliary words, connectives, interjections and games and routines. The Catalan inventory contained 778 items (196 cognates, 582 non-cognates) and the Spanish inventory contained 781 (197 cognates, 584 non-cognates).

@@

We gathered 181587 responses. Translation equivalents received an average of 386.36 (\emph{SD} = 228.76, , Max = 698), both languages summed together. We modelled the probability of responses (\(y_i\)) to belong to any of the three possible categories (\(k\)): 1 = \emph{No} or say, 2 = \emph{Understands}, and 3 = \emph{Understands and Says}) using an ordinal regression model with a cumulative logit link function (Agresti, 2010; Bürkner \& Vuorre, 2019). This model estimates the probability of response \(y_i\) to belong to category \(k\) or lower. In our case, this will result in the estimation of the probability of \(y_i \leq 1\), \(y_i \leq 2\), and \(y_i \leq 3\), where

@@

We implemented the model using the R package \texttt{brms} (Bürkner, 2017). This package uses the Stan probabilistic language (Carpenter et al., 2017) to approximate the posterior distribution of the model using Hamiltonian Monte Carlo. We implemented our extended model as \texttt{response\ \textasciitilde{}\ \ age\ +\ frequency\ +\ dominance*bilingualism*cognate}. We fit this model running four MCMC chains with 2,000 iterations and .80 delta parameter. We used weakly regularising prior for the parameters in our model, given that its weight in the estimation of the posterior will be marginal relative to that of the likelihood, due to the amount of observations introduced in the model. We then fit increasingly simpler models dropping predictors one by one in the following order: (1) \texttt{dominance:bilingualism:cognate}, (2) \texttt{bilingualism:cognate}, (3) \texttt{dominance:cognate:}, (4) \texttt{cognate}, (5) \texttt{dominance:bilingualism}, (6) \texttt{bilingualism}, (7) \texttt{dominance}, (8) \texttt{frequency}, and (9) \texttt{age}. We compared all models using leave-one-out cross-validation (LOO-CV) (Vehtari, Gelman, \& Gabry, 2017) and widely applicable information criterion (WAIC) (Watanabe \& Opper, 2010). We then explored the posterior distribution of each parameter in the model that fitted the data the best, computing the 95\% credible intervals and testing whether this interval excluded 0. We performed follow-up tests on interactions whose 95\% credible interval excluded 0 by comparing the 95\% credible interval of their estimated marginal means using the \texttt{emmeans} package (Lenth, 2020). Data processing and visualisation was done in R (R Core Team, 2020) using the \texttt{tidyverse} family of packages (Wickham et al., 2019), and posterior samples were extracted using the \texttt{tidybayes} R package (Kay, 2020).

@@

\leavevmode\hypertarget{ref-bosma2020cognate}{}%
Bosma, E., \& Nota, N. (2020). Cognate facilitation in frisian--dutch bilingual children's sentence reading: An eye-tracking study. \emph{Journal of Experimental Child Psychology}, \emph{189}, 104699.

@@

\leavevmode\hypertarget{ref-carpenter_stan_2017}{}%
Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B., Betancourt, M., \ldots{} Riddell, A. (2017). Stan: A Probabilistic Programming Language. \emph{Journal of Statistical Software}, \emph{76}(1), 1--32. \url{https://doi.org/10.18637/jss.v076.i01}

@@

\leavevmode\hypertarget{ref-costa2005facilitatory}{}%
Costa, A., Santesteban, M., \& Caño, A. (2005). On the facilitatory effects of cognate words in bilingual speech production. \emph{Brain and Language}, \emph{94}(1), 94--103.

@@

\leavevmode\hypertarget{ref-dell1991mediated}{}%
Dell, G. S., \& O'Seaghdha, P. G. (1991). Mediated and convergent lexical priming in language production: A comment on levelt et al (1991).

@@

\leavevmode\hypertarget{ref-doyle1977some}{}%
Doyle, A. B., \& others. (1977). Some issues in the assessment of linguistic consequences of early bilingualism. \emph{Working Papers on Bilingualism}, (14).

\leavevmode\hypertarget{ref-fernandez1992bilingual}{}%
Fernandez, M. C., Pearson, B. Z., Umbel, V. M., Oiler, D., \& Molinet-Molina, M. (1992). Bilingual receptive vocabulary in hispanic preschool children. \emph{Hispanic Journal of Behavioral Sciences}, \emph{14}(2), 268--276.

@@

@@

\leavevmode\hypertarget{ref-jardak2019labels}{}%
Jardak, A., \& Byers-Heinlein, K. (2019). Labels or concepts? The development of semantic networks in bilingual two-year-olds. \emph{Child Development}, \emph{90}(2), e212--e229.

@@

@@

\leavevmode\hypertarget{ref-poulin2013lexical}{}%
Poulin-Dubois, D., Bialystok, E., Blaye, A., Polonia, A., \& Yott, J. (2013). Lexical access and vocabulary development in very young bilinguals. \emph{International Journal of Bilingualism}, \emph{17}(1), 57--70.

@@

\leavevmode\hypertarget{ref-rosenblum1983word}{}%
Rosenblum, T., \& Pinker, S. A. (1983). Word magic revisited: Monolingual and bilingual children's understanding of the word-object relationship. \emph{Child Development}, 773--780. \url{https://doi.org/10.2307/1130064}

\leavevmode\hypertarget{ref-schelletter2002effect}{}%
Schelletter, C. (2002). The effect of form similarity on bilingual children's lexical development. \emph{Bilingualism: Language and Cognition}, \emph{5}(2), 93--107.

\leavevmode\hypertarget{ref-singh2014one}{}%
Singh, L. (2014). One world, two languages: Cross-language semantic priming in bilingual toddlers. \emph{Child Development}, \emph{85}(2), 755--766.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The role of cross-linguistic lexical similarity on bilingual word acquisition},
  pdfauthor={Gonzalo Garcia-Castro1, Daniela Avila-Varela1, \& Nuria Sebastian-Galles1},
  pdflang={en-EN},
  pdfkeywords={lexical acquisition, vocabulary, bilingualism},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\title{The role of cross-linguistic lexical similarity on bilingual word acquisition}
\author{Gonzalo Garcia-Castro\textsuperscript{1}, Daniela Avila-Varela\textsuperscript{1}, \& Nuria Sebastian-Galles\textsuperscript{1}}
\date{}

@@

Correspondence concerning this article should be addressed to Gonzalo Garcia-Castro, Ramon Trias Fargas, 25-27, 08005 Barcelona, Spain. E-mail: \href{mailto:gonzalo.garciadecastro@upf.edu}{\nolinkurl{gonzalo.garciadecastro@upf.edu}}

@@

The structure of the lexicon also impacts the order of acquisition of new words. Words that share a high degree of phonological and/or semantic overlap with words already acquired are more likely to be acquired next (Hills, Maouene, Maouene, Sheya, \& Smith, 2009). ({\textbf{???}}) and ({\textbf{???}}) showed that the connectivity of a given semantic category (i.e., animals) in a child's lexicon predicts a better performance in a disambiguation task, where participants are presented with a novel label in the presence of a familiar and a novel object: infants show stronger looking preference for the novel object if it belonged to a category for which many words had already been acquired. This points to the structure of the lexicon facilitating the strategies children engage during word learning.

@@

When compared to their monolingual peers, they know fewer words only when just on language is considered (e.g., English monolinguals know more words in English than English-Spanish bilinguals). When both languages are taken into account, bilingual children seem to know, at least, as many words as monolinguals do ({\textbf{???}}; {\textbf{???}}; Ben-Zeev, 1977; Bialystok, Luk, Peets, \& Yang, 2010; Blom et al., 2019; Core, Hoff, Rumiche, \& Señor, 2013; Fernandez, Pearson, Umbel, Oiler, \& Molinet-Molina, 1992; Hoff et al., 2012; Hoff, Rumiche, Burridge, Ribot, \& Welsh, 2014; Pearson \& Fernández, 1994; but see Pearson, Fernández, \& Oller, 1993; Houwer, Bornstein, \& Putnick, 2014).

Floccia et al. (2018), gathered parental vocabulary estimates (Hamilton, Plunkett, \& Schafer, 2000) from a large sample of 24-month-old toddlers learning English and an additional language (from a pool of 13 linguistically diverse languages). They computed the average phonological similarity between translation equivalents (TEs from now on; e.g., \emph{table} in English and \emph{tafel} in Dutch) for each pair of languages, and found a positive association between this measure of phonological similarity and participants' productive vocabulary sizes in their additional language. Using a similar measure of lexical similarity, Blom et al. (2019) extended these results to bilingual children aged three to 10, and reported a positive association between the lexical distance and comprehensive vocabulary size: bilinguals learning lexically close languages showed similar vocabulary sizes to those of monolinguals, but those learning the more distant languages showed lower vocabulary sizes.

These results suggest that linguistic distance between languages plays an important role during bilingual lexical acquisition, and that this effect operates within TEs. What mechanisms underlie this effect, and why they seem to operate differently across comprehension and production, or across the dominant and the non-dominant language, remain open issues. Floccia et al. (2018) pointed to parallel activation as a candidate mechanism behind this facilitation effect. The parallel activation principle suggests that bilinguals activate lexical representations in a language non-selective way: during the comprehension or production of a word in one language (e.g., \emph{cat}, in English) its corresponding translation in the other language is activated too {[}e.g., \emph{gato}, in Spanish. There is a vast body of evidence supporting language non-selective lexical access in both adults (e.g., Dell \& O'Seaghdha, 1991; {\textbf{???}}; Bobb, Von Holzen, Mayor, Mani, \& Carreiras, 2020; Costa \& Caramazza, 2000; Dijkstra, 2005; Hoshino \& Kroll, 2008; Kroll, Gullifer, \& Rossi, 2013; Singh, 2014; Thierry \& Wu, 2007; Von Holzen, Fennell, \& Mani, 2019; Yudes, Macizo, \& Bajo, 2010; see Costa, Santesteban, \& Caño, 2005 for review), and children (e.g.~Von Holzen \& Mani, 2012; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}). A critical implication of parallel activation is that the word form (i.e., phonology, orthography, etc.) of both translation equivalents are available to each other, impacting the comprehension or production dynamic of any of them.

@@

Under this account, cognates (i.e., phonologically similar TEs) would be acquired earlier than non-cognates, but this effect would only play a role once one of the members of the TE has been acquired. The evidence supporting an earlier age of acquisition of cognates relative to non-cognates if sparse. Schelletter (2002) reported a longitudinal case-study of a German/English child who produced cognate TEs earlier than non-cognate TE. Bosch and Ramon-Casas (2014), showed converging results from a sample of 48 monolingual and bilingual infants aged 24 months. The low number of words included in the analyses of both studies, and the lack of adjustment for lexical frequency, limits severely the strength of any conclusions concerning the difference in age of acquisition of cognates and non-cognates. On the other hand Floccia et al. (2018) worked with aggregated estimates of participants' vocabulary size, losing item-level information that could provide information about change in age of acquisition of TEs associated with their degree of phonological similarity. Therefore, their results can only be considered as compatible with cognateness playing a role on age of acquisition. In this study, we aimed at overcoming these pitfalls by using unaggregated data from individual responses, testing the role of cognateness explicitly and adjusting for lexical frequency.

@@

We implemented an on-line questionnaire using \texttt{formr} (Arslan, Walther, \& Tata, 2020), divided in three forms: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory (Hamilton et al., 2000) and consisted in two lists of words, one in Catalan and one in Spanish). Items in one language were translation equivalents of the items in the other (e.g., whenever \emph{gos} {[}dog{]} was included in the Catalan inventory, the word \emph{perro} was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both (e.g., Catalan \emph{acabar} {[}\emph{to finish}{]} and Spanish \emph{acabar} and \emph{terminar}), or merged them into the same one (e.g., Spanish \emph{mono} {[}\emph{monkey}{]} and Catalan \emph{mono/mico}). We included items from a diverse sample of 26 semantic/functional categories (see Appendix 1). We discarded the following categories: adverbs, auxiliary words, connectives, interjections and games and routines. The Catalan inventory contained 778 items (196 cognates, 582 non-cognates) and the Spanish inventory contained 781 (197 cognates, 584 non-cognates).

@@

We gathered 181587 responses. Translation equivalents received an average of 386.36 (\emph{SD} = 228.76, , Max = 698), both languages summed together. We modelled the probability of responses (\(y_i\)) belonging to any of the three possible categories \(k\)--1 = \emph{No} or say, 2 = \emph{Understands}, and 3 = \emph{Understands and Says}-- using an ordinal regression model with a cumulative logit link function (Agresti, 2010; Bürkner \& Vuorre, 2019). This model estimates the probability of response \(y_i\) to belong to category \(k\) or lower. In our case, this will result in the estimation of the probability of \(y_i \leq 1\), \(y_i \leq 2\), and \(y_i \leq 3\), where

@@

We implemented the model using the R package \texttt{brms} (Bürkner, 2017) as \texttt{response\ \textasciitilde{}\ \ age\ +\ frequency\ +\ dominance*bilingualism*cognate}, running four MCMC sampling chains with 4,000 iterations (including 2,000 warm-up iterations per chain). We used weakly regularising prior for the parameters in our model, given that its weight in the estimation of the posterior will be marginal due to the large amount of observations introduced in the model. We report the 95\% credible intervals (CrI) of the estimated posterior distribution of each parameter, which denotes a range of values we are 95\% certain contain the true value, given our prior and observed data. We report analyses on the probability scale, having transformed the log-odds returned by the model using an inverse logit transformation.

We fit increasingly simpler models dropping predictors one by one in the following order: (1) \texttt{dominance:bilingualism:cognate}, (2) \texttt{bilingualism:cognate}, (3) \texttt{dominance:cognate:}, (4) \texttt{cognate}, (5) \texttt{dominance:bilingualism}, (6) \texttt{bilingualism}, (7) \texttt{dominance}, (8) \texttt{frequency}, and (9) \texttt{age}. We compared all models using leave-one-out cross-validation (LOO-CV) (Vehtari, Gelman, \& Gabry, 2017)\footnote{Due to the high computational cost associated with a large dataset, we used a sub-sampling approach for performing Bayesian LOO with 1000 samples (Magnusson, Andersen, Jonasson, \& Vehtari, 2019)}. We then explored the posterior distribution of each parameter in the model that fitted the data the best, computing the 95\% credible intervals and testing whether this interval excluded 0. We performed follow-up tests on interactions whose 95\% credible interval excluded 0 by comparing the 95\% credible interval of their estimated marginal means using the \texttt{emmeans} package (Lenth, 2020). Data processing and visualisation was done in R (R Core Team, 2020) using the \texttt{tidyverse} family of packages (Wickham et al., 2019), and posterior samples were extracted using the \texttt{tidybayes} R package (Kay, 2020).

\hypertarget{results}{%
\section{Results}\label{results}}

Table 2 summarises shows the outcome of the LOO-CV model comparison. The extended model was most favoured by the data.

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{crrrrr}
\toprule
& & & & \multicolumn{2}{c}{\textbf{Comparison}} \\ 
 \cmidrule(lr){5-6}
\textbf{Model} & \textbf{ELPD}\textsuperscript{a} & \textbf{SE}\textsuperscript{b} & \textbf{p}\textsuperscript{c} & \emph{LOOdiff} & \emph{SEdiff} \\ 
\midrule
\textasciitilde  1 (intercept only) & $-157,272.37$ & $250.56$ & $12.00$ & $0.00$ & 0.0000 \\ 
... + age & $-157,308.76$ & $250.47$ & $10.37$ & $36.39$ & 354.2836 \\ 
... + frequency & $-157,318.93$ & $250.45$ & $9.93$ & $46.56$ & 354.2703 \\ 
... + dominance & $-157,463.31$ & $250.25$ & $8.08$ & $190.94$ & 354.1293 \\ 
... + bilingualism & $-157,472.54$ & $250.29$ & $7.67$ & $200.17$ & 354.1564 \\ 
... + dominance:bilingualism & $-158,724.35$ & $246.22$ & $6.33$ & $1,451.98$ & 351.2863 \\ 
... + cognate & $-159,063.42$ & $245.10$ & $5.09$ & $1,791.06$ & 350.5085 \\ 
... + dominance:cognate & $-161,704.16$ & $237.62$ & $3.95$ & $4,431.79$ & 345.3184 \\ 
... + bilingualism:cognate & $-165,570.87$ & $228.08$ & $3.13$ & $8,298.50$ & 338.8223 \\ 
... + dominance:bilingualism:cognate & $-184,511.91$ & $172.86$ & $2.04$ & $27,239.54$ & 304.4033 \\ 
\bottomrule
\end{longtable}
\vspace{-5mm}
\begin{minipage}{\linewidth}
\textsuperscript{a}ELPD: theoretical expected log pointwise predictive density for a new dataset, as estimated using Bayesian LOO-CV \\ 
\textsuperscript{b}SE: standard error of the ELPD \\ 
\textsuperscript{c}p: effective number of parameters \\ 
\end{minipage}

Table 3 shows the mean and 95\% CrI of each regression coefficient of the best fitting model.

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{\textbf{Odds Ratio}} & \multicolumn{3}{c}{\textbf{Probability}} \\ 
 \cmidrule(lr){2-4}\cmidrule(lr){5-7}
\textbf{Coefficient} & \emph{Mean} & \emph{SE}\textsuperscript{1} & \emph{95\% CrI}\textsuperscript{2} & \emph{Mean} & \emph{SE} & \emph{95\% CrI} \\ 
\midrule
Intercept 1 (Understands) & $1.12$ & $1.01$ & $1.11$–$1.14$ & $0.53$ & $0.50$ & $0.53$–$0.53$ \\ 
Intercept 2 (Understands \& Says) & $5.31$ & $1.01$ & $5.24$–$5.38$ & $0.84$ & $0.50$ & $0.84$–$0.84$ \\ 
Age & $1.16$ & $1.00$ & $1.16$–$1.17$ & $0.54$ & $0.50$ & $0.54$–$0.54$ \\ 
Frequency & $1.86$ & $1.01$ & $1.83$–$1.88$ & $0.65$ & $0.50$ & $0.65$–$0.65$ \\ 
Dominance & $2.27$ & $1.01$ & $2.22$–$2.32$ & $0.69$ & $0.50$ & $0.69$–$0.70$ \\ 
Bilingualism & $1.10$ & $1.00$ & $1.09$–$1.11$ & $0.52$ & $0.50$ & $0.52$–$0.53$ \\ 
Cognate & $1.08$ & $1.01$ & $1.06$–$1.10$ & $0.52$ & $0.50$ & $0.51$–$0.52$ \\ 
Dominance * Bilingualism & $0.73$ & $1.01$ & $0.73$–$0.74$ & $0.42$ & $0.50$ & $0.42$–$0.43$ \\ 
Dominance * Cognate & $0.68$ & $1.02$ & $0.65$–$0.71$ & $0.40$ & $0.51$ & $0.39$–$0.42$ \\ 
Bilingualism * Cognate & $0.97$ & $1.01$ & $0.96$–$0.98$ & $0.49$ & $0.50$ & $0.49$–$0.50$ \\ 
Dominance * Bilingualism * Cognate & $1.13$ & $1.01$ & $1.10$–$1.15$ & $0.53$ & $0.50$ & $0.52$–$0.54$ \\ 
\bottomrule
\end{longtable}
\vspace{-5mm}
\begin{minipage}{\linewidth}
\textsuperscript{1}SE: Standard Error of the mean \\ 
\textsuperscript{2}95\% CrI: 95\% credible interval \\ 
\end{minipage}

\begin{figure}

{\centering \includegraphics[width=20cm]{trajectories_manuscript_files/figure-latex/marginalmeans-1}

\caption{Mean posterior probability and 95\% CrI of responses, conditional to Dominance, Bilingualism, and Cognate.}\label{fig:marginalmeans}
\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

@@

@@

\leavevmode\hypertarget{ref-costa_lexical_2000}{}%
Costa, A., \& Caramazza, A. (2000). Lexical access in speech production: The bilingual case. \emph{Psicológica}, \emph{21}(2), 403--437.

\leavevmode\hypertarget{ref-costa_facilitatory_2005}{}%
Costa, A., Santesteban, M., \& Caño, A. (2005). On the facilitatory effects of cognate words in bilingual speech production. \emph{Brain and Language}, \emph{94}(1), 94--103.

@@

\leavevmode\hypertarget{ref-dell_mediated_1991}{}%
Dell, G. S., \& O'Seaghdha, P. G. (1991). Mediated and convergent lexical priming in language production: A comment on levelt et al (1991).

@@

\leavevmode\hypertarget{ref-fernandez_bilingual_1992}{}%
Fernandez, M. C., Pearson, B. Z., Umbel, V. M., Oiler, D. K., \& Molinet-Molina, M. (1992). Bilingual Receptive Vocabulary in Hispanic Preschool Children. \emph{Hispanic Journal of Behavioral Sciences}, \emph{14}(2), 268--276. \url{https://doi.org/10.1177/07399863920142006}

@@

\leavevmode\hypertarget{ref-hills_longitudinal_2009}{}%
Hills, T. T., Maouene, M., Maouene, J., Sheya, A., \& Smith, L. (2009). Longitudinal Analysis of Early Semantic Networks Preferential Attachment or Preferential Acquisition? \emph{Psychological Science}, \emph{20}(6), 729--739. \url{https://doi.org/10.1111/j.1467-9280.2009.02365.x}

@@

@@

\leavevmode\hypertarget{ref-magnusson_leave_2019}{}%
Magnusson, M., Andersen, M., Jonasson, J., \& Vehtari, A. (2019). Bayesian leave-one-out cross-validation for large data. In K. Chaudhuri \& R. Salakhutdinov (Eds.), \emph{Proceedings of the 36th international conference on machine learning} (Vol. 97, pp. 4244--4253). Long Beach, California, USA: PMLR. Retrieved from \url{http://proceedings.mlr.press/v97/magnusson19a.html}

@@

@@

\leavevmode\hypertarget{ref-schelletter_effect_2002}{}%
Schelletter, C. (2002). The effect of form similarity on bilingual children's lexical development. \emph{Bilingualism: Language and Cognition}, \emph{5}(2), 93--107. \url{https://doi.org/10.1017/S1366728902000214}

\leavevmode\hypertarget{ref-singh_one_2014}{}%
Singh, L. (2014). One world, two languages: Cross-language semantic priming in bilingual toddlers. \emph{Child Development}, \emph{85}(2), 755--766.
